---
title: "Coursera ML Week4: Practical Machine Learning"
author: "Deepesh Madkar"
date: "8/19/2018"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<hr>
#### Overview:
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

#### Links For The Raw Data:

<p>
The data for this project are available here:
<ul>
  <li>
  Link to download traning data: [Training Data CSV]("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
  </li>
  <li>
  Link to download test data: [Test Data CSV]("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv") 
  </li> 
  <li>
   Link to main source: [Main Source ]("http://groupware.les.inf.puc-rio.br/har") 
  </li>
</ul>
</p>

```{r}
# To reproduce the result please change the working directory
workingDirectory = "/Projects/Rworkspace/coursera/courseraMLWeek4"
setwd(workingDirectory)
```
#### Loading data & performing some basic exploratory data analysis:
```{r results='hide', message=FALSE, warning=FALSE}
# Packages
library(caret)
```


```{r cache = TRUE}

# checking if data directory exists if not creating it
if(!file.exists("./data")){
  dir.create("data")
  
  # Downloading the file 
  trainDataUrlPath <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(trainDataUrlPath, "./data/trainData.csv", method="curl")
  
  # Downloading the file 
  testDataUrlPath <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(testDataUrlPath, "./data/testData.csv", method="curl")
}
```

Taking the overview of the training data before processing.
</i></p>
```{r cache = TRUE}
# Loading the dataset
trainDataset <- read.csv("./data/trainData.csv", header = TRUE, sep = ",", quote = "\"", na.strings=c("NA","#DIV/0!",""))
validationDataset <- read.csv("./data/testData.csv", header = TRUE, sep = ",", quote = "\"", na.strings=c("NA","#DIV/0!",""))

# Cleaning the data
# removing near zero values
nearZeroVar <- nearZeroVar(trainDataset)
trainDataset <- trainDataset[, -nearZeroVar]

# removing na's 
nalVal <- sapply(trainDataset, function(x) mean(is.na(x))) > 0.95
trainDataset <- trainDataset[, nalVal==FALSE]

# removing the column 1:5
trainDataset <- trainDataset[, -c(1:5)]

# Partitioning the data
inTrain  <- createDataPartition(trainDataset$classe, p=0.7, list=FALSE)
trainingDataset <- trainDataset[inTrain,]
testingDataset <- trainDataset[-inTrain,]

# Checking the dimmension of the training data
dim(trainingDataset)

# Checking the dimmension of the validation data
dim(testingDataset)

```
#### Model Building
####Using Decision Tree - Classification Model:
```{r cache = TRUE}
# setting seed
set.seed(987)

# Fitting model using caret package - (method = rpart, i.e: decision tree)
modelFitDecionTree <- train(classe ~ ., data=trainingDataset, method="rpart")
modelFitDecionTree

# Predicting on the test set
modelPredictDecionTree <- predict(modelFitDecionTree, newdata=testingDataset)
compDecisionTree <- confusionMatrix(modelPredictDecionTree, testingDataset$classe)
compDecisionTree
```
<i>The following are the results obtained by Decision Tree model</i>
<ul>
  <li>Decision Tree Accuracy: `r compDecisionTree$overall['Accuracy']` </li>
  <li>The out-of-sample error: `r 1 - compDecisionTree$overall['Accuracy']`</li>
</ul>

```{r cache = TRUE}
# graphical overview
plot(compDecisionTree$table, col = compDecisionTree$byClass)
```

<hr>
#### Using Random Forest:
```{r cache = TRUE}
# setting seed
set.seed(657)

# defining cross validation parameter for the model
crossValidationParamRF <- trainControl(
    method="cv",
    number=5,
    savePredictions = TRUE,
    classProbs = TRUE,
    verboseIter=FALSE
    )

# Fitting model using caret package - (method = rf, i.e: random forest)
modelFitRandomForest <- train(classe ~ ., data=trainingDataset, method="rf",
                          trControl=crossValidationParamRF)
modelFitRandomForest

# Predicting on the test set
modelPredictRandomForest <- predict(modelFitRandomForest, newdata=testingDataset)

compRandomForest <- confusionMatrix(modelPredictRandomForest, testingDataset$classe)
compRandomForest
```
<i>The following are the results obtained by Random Forest</i>
<ul>
  <li>Random Forest Accuracy: `r compRandomForest$overall['Accuracy']` </li>
  <li>The out-of-sample error: `r 1 - compRandomForest$overall['Accuracy']`</li>
</ul>

```{r cache = TRUE}
# graphical overview
plot(compRandomForest$table, col = compRandomForest$byClass)
```

<hr>
#### Using Generalized Boosted Model:
```{r cache = TRUE, results='hide'}
# setting seed
set.seed(345)

# defining cross validation parameter for the Generalized Boosted Regression Model
crossValidationParamGB <- trainControl(
    method="repeatedcv",
    number=5,
    repeats = 1,
    savePredictions = TRUE,
    classProbs = TRUE
    )
# Fitting model using caret package - (method = gbm, i.e: generalized boosted regression)
modelFitGenBoost <- train(classe ~ ., data=trainingDataset, method="gbm",
                          trControl=crossValidationParamGB)


# because it has huge output, i have commented it
# modelFitGenBoost
```

```{r cache = TRUE, results='hide'}
# Predicting on the test set
modelPredictGenBoost <- predict(modelFitGenBoost, newdata=testingDataset)

compGenBoost <- confusionMatrix(modelPredictGenBoost, testingDataset$classe)
compGenBoost

```
<i>The following are the results obtained by Generalized boosted model</i>
<ul>
  <li>Generalized boosted model Accuracy: `r compGenBoost$overall['Accuracy']` </li>
  <li>The out-of-sample error: `r 1 - compGenBoost$overall['Accuracy']`</li>
</ul>

```{r cache = TRUE}
# graphical overview
plot(compGenBoost$table, col = compGenBoost$byClass)
```

<hr>
#### Model Selection/Conclusion:
So based on the accuracy, the results of random forest model are more accurate than the other models. so applying the results to the validation dataset.

```{r cache = TRUE}
# Predicting on the validation Dataset
finalPrediction <- predict(modelFitRandomForest, newdata=validationDataset)
finalPrediction
```


